---
title: "Comparing the speed of graph drawing in ggplot2 and base plot"
author: "Samantha Tyner, Francoise Briatte, and Heike Hofmann"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE}

knitr::opts_chunk$set(fig.width = 6,
                      fig.height = 4,
                      fig.align='center',
                      dev = "png")

```

# Introduction

This vignette compares the speeds of our package to the other `ggplot2`-based network visualization package `ggnetwork`, the `GGally` function `ggnet2()`, and the visualization capabilities of the `igraph` and `network` packages. 

# Speed Comparison 1 

In this first example, we visualize the protein-protein interaction network in the yeast species *S. cerevisiae* using several different methods.

First, load all of the necessary packages and data sets.

```{r speed1_loadpkg, echo = TRUE, eval = TRUE, message=FALSE, warning = FALSE}
## igraph (current: v1.0.1)
if (!require(igraph, quietly = TRUE)) {
  install.package("igraph")
}
library(igraph)

## network (current: v1.13.0)
if (!require(network, quietly = TRUE)) {
  install.package("network")
}
library(network)

## ggplot2
if (!require(ggnetwork, quietly = TRUE) ||
    packageVersion("ggplot2") < "1.0.1.9003") {
  devtools::install_github("hadley/ggplot2")
}
library(ggplot2)

## ggnet2
if (!require(GGally, quietly = TRUE)) {
  install.packages("GGally")
}
library(GGally)

## geom_net
if (!require(geomnet, quietly = TRUE)) {
  devtools::install_github("sctyner/geomnet")
}
library(geomnet) # also currently requires dplyr

## ggnetwork
if (!require(ggnetwork, quietly = TRUE)) {
  install.packages("ggnetwork")
}
library(ggnetwork)

## other packages
library(dplyr)
library(tidyr)

## pre-loading
library(scales)
library(sna)

## load data 
data(protein, package = "geomnet")
```

Next, generate the network visualization of this network 100 times using the random layout option in each package and store the time to plot each run. 

```{r speed1_simcode, eval=FALSE, echo=TRUE}
d = data.frame()

if (!file.exists("runtimes-protein-100.csv")) {
  
  for (i in 1:100) {
  
    cat("Iteration", sprintf("%3.0f", i), "/ 100\n")
  
    n = as.matrix(protein$edges[, 1:2])
    n = igraph::graph_from_edgelist(n, directed = FALSE)
  
    t1 = system.time({
      plot(n, vertex.label = NA, layout = layout_randomly)
    })[1]
  
    n = network(protein$edges[, 1:2], directed = FALSE)

    t2 = system.time({
     plot(n, coord = gplot.layout.random(n, NULL))
    })[1]

    t3 = system.time({
      print(ggnet2(n, mode = "random"))
    })[1]

    t4 = system.time({
      print(ggplot(data = protein$edges, aes(from_id = from, to_id = to)) +
              geom_net(layout = "random"))
    })[1]

    t5 = system.time({
      print(ggplot(ggnetwork(n, layout = "random"),
                   aes(x, y, xend = xend, yend = yend)) +
              geom_edges() +
              geom_nodes())
    })[1]
  
    d = rbind(d, data.frame(
      iteration = i,
      igraph = t1,
      network = t2,
      ggnet2 = t3,
      geomnet = t4,
      ggnetwork = t5,
      row.names = NULL
    ))

  }

  write.csv(d, file = "runtimes-protein-100.csv", row.names = FALSE)
  
}
```

Finally, compare the 100 runtimes for each method side-by-side. In the resulting plot below, we use our own data, which can be found [here](https://raw.githubusercontent.com/sctyner/ggnet-paper/master/runtimes-protein/runtimes-protein-100.csv). 

```{r speed1_cheatdata, echo=FALSE}
link_to_data <- "https://raw.githubusercontent.com/sctyner/ggnet-paper/master/runtimes-protein/runtimes-protein-100.csv"
```
```{r speed1_viewdata, echo = TRUE, eval = FALSE}
link_to_data <- "runtimes-protein-100.csv"
```
```{r speed1_plot, echo=TRUE}
g = read.csv(link_to_data, stringsAsFactors = FALSE) %>%
      gather(`Visualization approach`, time, -iteration)

ggplot(g, aes(x = `Visualization approach`, y = time,
              fill = `Visualization approach`)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  labs(y = "\nPlotting time (seconds)", x = "Visualization approach\n") +
  guides(fill = FALSE) +
  theme_classic(18) +
  theme(axis.text = element_text(size = rel(1)))
```

# Speed Comparison 2

For the second speed comparison, we tested the speeds of these same packages on 11 different machines (our own machines and some of our friends' and colleagues' machines). The data containing runtimes of for all of these machines can be found [here](https://github.com/sctyner/ggnet-paper/tree/master/runtimes). 

This time, we visualize graphs of various sizes that we generated randomly with edge probability 0.2 using the `sna` package. The sizes we generated are overall fairly small, from 25 to 250 nodes, going up by 25 each time. Each network visualization method was test 100 times in total for each size of network. 

```{r speed_2code, eval = FALSE}
# note: plot.network might be slower because it actually prints the plots, so to
# compare the methods, we need to evaluate the ggplot plots by printing them too

# create 10 files per size 
for (k in 0:9) {

  # the 10 different sizes   
  for (i in seq(250, 25, -25)) {
    
    f = paste0("runtimes-", sprintf("%04.0f", i), "-", k, ".csv")
    
    if (!file.exists(f)) {

      d = data.frame()

      # in each file store the runtimes of plotting 10 random graphs 
      # do this for each of the 5 methods 
      for (j in 1:10) {

        r = sna::rgraph(i, tprob = 0.2)
      
        cat("Network size", i,
            "iteration", sprintf("%3.0f", 10 * k + j), "/ 100\n")

        n = igraph::graph_from_adjacency_matrix(r, mode = "undirected")

        t1 = system.time({
          plot(n, vertex.label = NA)
        })[1]

        n = network::network(r, directed = FALSE)

        t2 = system.time({
          plot.network(n)
        })[1]

        t3 = system.time({
          print(ggnet2(n))
        })[1]

        e = data.frame(sna::as.edgelist.sna(n))

        t4 = system.time({
          print(ggplot(data = e) +
                  geom_net(aes(from_id = X1, to_id = X2)))
        })[1]

        t5 = system.time({
          print(ggplot(ggnetwork(n),
                       aes(x, y, xend = xend, yend = yend)) +
                  geom_edges() +
                  geom_nodes())
        })[1]

        d = rbind(d, data.frame(
          network_size = i,
          iteration = 10 * k + j,
          igraph = t1,
          network = t2,
          ggnet2 = t3,
          geomnet = t4,
          ggnetwork = t5,
          row.names = NULL
        ))

        dev.off()

      }

      write.csv(d, f, row.names = FALSE)

    }
        
  }
```

Again, for this plot we use our data that is available on Github.

```{r speed2_cheat, echo=FALSE}
  g = paste0("~/Desktop/NetworksResearch/ggnet-paper/runtimes/", list.files(path = "~/Desktop/NetworksResearch/ggnet-paper/runtimes", pattern = "runtimes-(.*)csv$", recursive = T)) %>%
        lapply(read.csv, stringsAsFactors = FALSE) %>%
        bind_rows %>%
        gather(`Visualization approach`, time, -network_size, -iteration) %>%
        group_by(network_size, `Visualization approach`) %>%
        summarise(mean_time = mean(time),
                  q05 = quantile(time, 0.05),
                  q95 = quantile(time, 0.95))
```
```{r speed2_code, echo = TRUE, eval=FALSE}  
# get all files containing runtime information 
g = list.files(pattern = "runtimes-(.*)csv$") %>%
        lapply(read.csv, stringsAsFactors = FALSE) %>%
        bind_rows %>%
        gather(`Visualization approach`, time, -network_size, -iteration) %>%
        group_by(network_size, `Visualization approach`) %>%
        summarise(mean_time = mean(time),
                  q05 = quantile(time, 0.05),
                  q95 = quantile(time, 0.95))
```

```{r speed2_plot}
   ggplot(g, aes(x = network_size, y = mean_time,
                    fill = `Visualization approach`)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 1/3) +
    geom_line(aes(color = `Visualization approach`)) +
    scale_y_continuous(breaks = seq(0, max(g$q95), 5)) +
    scale_x_continuous(breaks = seq(25, 250, 25)) +
    labs(y = "Average plotting time (seconds)\n", x = "\nNetwork size") +
    theme_classic(18) +
    theme(
      axis.text = element_text(size = rel(1)),
      legend.text = element_text(size = rel(1)),
      legend.justification = c(0,1),
      legend.position = c(0,1)
    )
```